A thermodynamic view of dusty protoplanetary disks: axisymmetric stability

### 2016-10-21 model.tex
tiny correction in entropy def
last update until november, for real

The gas component of the mixture has a specific entropy
$S = C_p\ln{P/\rhog^\gamma}$, where $C_p$ is the heat capacity at constant
pressure. However, this is not a useful quantity for (locally)
isothermal gas.

@@

In this section we review the basic properties of the 
single-fluid mixture using Eqs.\ref{masseq}--\ref{momeq} with 
Eq. \ref{eff_energy} in place of the dust continuity equation. Our
discussion is based on the corresponding axisymmetric linearized
equations. A variable $f$ is
subject to Eulerian perturbations of the form 
\begin{align}
 \real\left[ \delta f(r,z)\exp{\left(-\ii\sigma t\right)}\right], 
\end{align}
where $\sigma$ is the complex mode frequency. In Appendix
\ref{var_prin} we derive the integral relation,  
\begin{align}
%&  \sigma^2\int\rho\left(|\dd v_r|^2 + |\dd v_z|^2\right)dV \notag\\
  \sigma^2\mathcal{I}^2
&= \int\left[ \rho
  |\dd v_r|^2A + \rho  \dd v_z \dd v_r^* B + \rho \dd v_z^*\dd v_r C +
  \rho |\dd v_z|^2 D\phantom{\frac{1}{1}}\right. \notag\\
&\phantom{===}  \left. + \frac{1}{P}\Bigl\lvert \nabla\cdot\left(P\dd
  \bm{v}\right)\Bigr\rvert^2\right]dV  -\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV \notag\\
&\phantom{===}
- \int P
  \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV.\label{int_rel}
\end{align} 
The real integral $\mathcal{I}^2>0$ and coefficients $A,B,C,D$ can be
read off Eq. \ref{integral_ex}. %Note that $B=C$ from dynamical equilibrium
%(Eq. \ref{vshear}). 
We now consider various limits of Eq. \ref{int_rel}.

@@

\subsection{Thermodynamics of dust-drag instabilities}\label{dust_work}
Now consider constant $c_s$ but $\tstop\neq0$. 
Eq. \ref{int_rel} indicates that $\sigma^2$ is generally complex, so
we may have growing oscillations (overstability). Writing $\sigma =
-\omega + \ii s$ where $\omega$ and $s$ are the real frequency and
growth rates, respectively, the imaginary part of Eq. \ref{int_rel}
gives  
\begin{align}
%  s = \frac{\imag\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV}{2\omega\int\rho\left(|\dd
%    v_r|^2 + |\dd v_z|^2\right)dV}, \label{thermal_instability}
  s = \frac{\imag\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV}{2\omega\mathcal{I}^2}. \label{thermal_instability}
\end{align}
This represents instability due to dust-gas friction. 
Since drag ($\delta \mathcal{C}$) appears as a source term in our
effective energy equation, the quantity 
$\imag\left[\left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}\right]$
may be thought of as a correlation between compression/expansion and  
heating/cooling.

>>>>

The gas component of the mixture has a specific entropy
$S = C_p\ln{\left(P^{1/\gamma}/\rhog\right)}$, where $C_p$ is the heat capacity at constant
pressure. However, this is not a useful quantity for (locally)
isothermal gas.

@@

In this section we review the basic properties of the 
single-fluid mixture using Eqs.\ref{masseq}--\ref{momeq} with 
Eq. \ref{eff_energy} in place of the dust continuity equation. Our
discussion is based on the corresponding axisymmetric linearized
equations. A variable $f$ is
subject to Eulerian perturbations of the form 
\begin{align}
 \real\left[ \delta f(r,z)\exp{\left(-\ii\sigma t\right)}\right], 
\end{align}
where $\sigma$ is the complex mode frequency. For later convenience we
write 
\begin{align}
  \sigma = \ii s - \omega,
\end{align}
where $s$ and $\omega$ are the growth rate and real frequencies,
respectively. Thus perturbations have time dependence $e^{st +
  \ii\omega t}$.

In Appendix
\ref{var_prin} we derive the integral relation,  
\begin{align}
%&  \sigma^2\int\rho\left(|\dd v_r|^2 + |\dd v_z|^2\right)dV \notag\\
  \sigma^2\mathcal{I}^2
&= \int\left[ \rho
  |\dd v_r|^2A + \rho  \dd v_z \dd v_r^* B + \rho \dd v_z^*\dd v_r C +
  \rho |\dd v_z|^2 D\phantom{\frac{1}{1}}\right. \notag\\
&\phantom{===}  \left. + \frac{1}{P}\Bigl\lvert \nabla\cdot\left(P\dd
  \bm{v}\right)\Bigr\rvert^2\right]dV  -\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV \notag\\
&\phantom{===}
- \int P
  \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV.\label{int_rel}
\end{align} 
The real integral $\mathcal{I}^2>0$ and coefficients $A,B,C,D$ can be
read off Eq. \ref{integral_ex}. %Note that $B=C$ from dynamical equilibrium
%(Eq. \ref{vshear}). 
We now consider various limits of Eq. \ref{int_rel}.

@@

\subsection{Thermodynamics of dust-drag instabilities}\label{dust_work}
Now consider constant $c_s$ but $\tstop\neq0$. 
Eq. \ref{int_rel} indicates that $\sigma^2$ is generally complex, so
we may have growing oscillations (overstability). %Writing $\sigma =
%-\omega + \ii s$ where $\omega$ and $s$ are the real frequency and
%growth rates, respectively, 
The imaginary part of Eq. \ref{int_rel}
gives  
\begin{align}
%  s = \frac{\imag\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV}{2\omega\int\rho\left(|\dd
%    v_r|^2 + |\dd v_z|^2\right)dV}, \label{thermal_instability}
  s = \frac{\imag\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV}{2\omega\mathcal{I}^2}. \label{thermal_instability}
\end{align} 
This represents instability due to dust-gas friction. 
Since drag ($\delta \mathcal{C}$) appears as a source term in our
effective energy equation, the quantity 
$\imag\left[\left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}\right]$
may be thought of as a correlation between compression/expansion and  
heating/cooling.
